\pkg{dataMaid}: your assistant for data cleaning in \proglang{R}

### 2017-01-24 latex/article_vol2.Rnw
Chunks

<<echo=FALSE>>=
options(prompt = "R> ", width=77, continue = "+  ")
@

@@

Online tools such as OpenRefine (\url{http://openrefine.org/}) and
\proglang{R}-packages such as \pkg{plyr} \citep{plyr}, and \pkg{data.table} have
made data wrangling a lot easier, but only a handful of packages such
as \pkg{editrules}, \pkg{validate}, \pkg{DataCombine}, and
\pkg{janitor} attempt to implement systematic, reproducible data
cleaning.  These packages use different approaches for data cleaning:
\pkg{editrules} and \pkg{validate} provide frameworks for setting up
and checking constraints on the variables, while \pkg{DataCombine} and
\pkg{janitor} both provide a few functions for identifying problems
(e.g, duplicates, dates coded as numbers, etc.) in data.

While these tools attempt to alleviate the ubiquitous ad hoc approach
to data cleaning, they are primarily intended for the data savvy users
and less so for the general researcher with a knowledge about a specific
field and the context of the available data. The \pkg{dataMaid} package
tries to address this by providing a framework that both allows for
extendable, systematic, reproducible data cleaning, and summarizing
findings for researchers from other fields such that they can act as
human experts when tracking down potential errors.

@@

The \pkg{dataMaid} approach to data cleaning and -quality assessment is
governed by two fundamental paradigms. First of all, there is no need
for data cleaning to be an ad hoc procedure. Often, we have a very
clear idea of what flags are raisable in a given dataset before we
look at it, as we were the ones to produce it in the first place. This
means that data cleaning can easily be a well-documented,
well-specified procedure. In order to aid this paradigm, \pkg{dataMaid}
provides easy-to-use, automated tools for data quality assessment in
\proglang{R} on which data cleaning decisions can be made. This quality
assessment is presented in an auto-generated overview document,
readable by data analysts and field experts alike, thereby also
contributing to an inter-field dialogue about the data at
hand. Oftentimes, e.g., distinguishing between faulty codings of a
numeric value and unusual, but correct, values requires
problem-specific expertise that might not be held by the data
analyst. Hopefully, having easy access to data descriptions through
\pkg{dataMaid} will help this necessary knowledge sharing.

@@

The final rendering of the generated markdown file is controlled by
the \code{render} and \code{openResult} arguments, which both default to
\code{TRUE}. \code{render} determines if the \proglang{R} markdown file
produced should be rendered using the \pkg{rmarkdown} package and
\code{openResult} decides whether the output html or pdf file should be
opened. The following command produces an \proglang{R} markdown file
containing the data cleaning summary information, but without
rendering nor opening the markdown file:

@@

Note that the data visualizations in the cleaning summary are also
controllable, though only a single function can be provided for all
variable types. If for instance, we wish to change the visualizations
from the default \pkg{ggplot2} style histograms and barplots to base
\proglang{R} histograms and barplots, we type

@@

\begin{table}[htbp]
\footnotesize
\bgroup
\def\arraystretch{1.8}%
\begin{tabular}{L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}}
& \code{summaryFunction} & \code{visualFunction} & \code{checkFunction} \\
\hline
\vspace{0pt} Input (required) & \code{v} - a variable vector \newline \code{...}  &  \code{v} - a variable vector \newline \code{vnam} - the variable name (as character string) \newline \code{doEval} - a logical (\code{TRUE}/\code{FALSE}) controlling the output type of the function & \code{v} - a variable vector \newline \code{nMax} - an integer (or \code{Inf}), controlling how many problematic values are printed, if relevant \newline \code{...}  \\
Input (optional) &  \code{maxDecimals} - number of decimals printed in outputted numerical values.  & - &  \code{maxDecimals}  - number of decimals printed in outputted numerical values.  \\
Purpose & Describe some aspect of the variable, e.g., a central value, its dispersion or level of missingness. & Produce a distribution plot. & Check a variable for a specific issue and, if relevant, identify the values in the variable that cause the issue. \\
Output (required) & A list with entries: \newline \code{\$feature} - a label for the summary value (as character string); \newline \code{\$result} - the result of the summary (as character string) & A character string with \code{R} code for producing a plot. This code should be standalone, i.e., should include the data if necessary. & A list with entries: \newline \code{\$problem} - a logical identifying whether an issue was found; \newline \code{\$message} - a character string (possibly empty) decribing the issue that was found, properly escaped and ready for use in \code{rmarkdown} \\
Output (recommended) & A \code{summaryResult} object, i.e., an attributed list with entries \code{\$feature}, \code{\$result} and \code{\$value}, the latter being the values from \code{\$result} in their original format). & \bgroup
\def\arraystretch{1.8}%
\hspace*{-.2cm}\begin{tabular}[t]{p{0.45\linewidth} |
                 p{0.45\linewidth}} \vspace*{-1cm}\raggedright If
                 \code{doEval} is \code{TRUE}: \newline A plot that is
                 opened by the graphic device in \code{R}. &
                                                          \vspace*{-1cm}\raggedright If \code{doEval} is \code{FALSE}: \newline   A text string with \code{R} code, as described above. \end{tabular} \egroup  & A \code{checkResult} object, i.e., an attributed list with entries \code{\$problem}, \code{\$message} and \code{\$problemValues}, the latter being either \code{NULL} or the problem causing values, as they were found in \code{v}, whichever is relevant. \\
Tools available for producing the function & \code{summaryResult()} & --- & \code{messageGenerator()} \newline \code{checkResult()}
\end{tabular}
\egroup
\caption{Reference information for creating new functions to be used
  as part of the \mbox{summarize-,} visualize-, and check steps. The three
  columns correspond to each of the three function types. %\hl{I think the formatting here is still sort of chaotic. Maybe include horizontal lines everywhere? Is it allowed in JSS? It's impossible to read what's in which row right now.}
}
\label{table.functionTypes}
\end{table}

@@

This function can now be called directly or used in \code{clean()}, as will
presented in an example below. Depending on the \code{doEval} argument, either a text string
with code or a plot is produced. The plot resulting from the following
call is found in Figure~\ref{fig:mosaicPlot}:
\begin{Schunk}
\begin{Sinput}
> mosaicVisual(toyData$var1, "variable 1", doEval = TRUE)
\end{Sinput}
\end{Schunk}
% $

@@

As with \code{mosaicVisual()}, we change the class of this function in
order to make it appear in \code{allSummaryFunctions()} calls. But now we
wish to emphasize that the function is not intended to be called on
all variable types, as zeros have different roles in \code{Date}s and in
\code{logical} variables:
\begin{Schunk}
\begin{Sinput}
> countZeros <- summaryFunction(countZeros,
	description = "Count number of zeros",
	classes = c("character", "factor", "integer",
                    "labelled", "numeric"))
\end{Sinput}
\end{Schunk}
% \hl{more? don't end on code.}

\subsubsection{identifyColons - a new checkFunction with problem values}
The last function mentioned above is \code{identifyColons()}. We define
it using the helper function \code{messageGenerator()} to obtain a properly
escaped message, and we use \code{checkResult} to make its output print
neatly:
\begin{Schunk}
\begin{Sinput}
identifyColons <- function(v, nMax = Inf, ... ) {
  v <- unique(na.omit(v))
  problemMessage <- "Note: The following values include colons:"
  problem <- FALSE
  problemValues <- NULL

@@

identifyColons <- checkFunction(identifyColons,
    description = "Identify colons surrounded by alphanumeric characters",
    classes = c("character", "factor", "labelled"))
\end{Sinput}
\end{Schunk}
As with the previous two functions, we also change its class. Note,
however, that for \code{checkFunction}s, the function description will
appear in the document produced by \code{clean()} (in the \textit{Data
  cleaning summary} section of the output), so now this is not only
done for the sake of the \code{allCheckFunctions()} output.

\subsubsection{Calling the new summarize/visualize/check functions from clean()}
Now, we are ready to use these new functions in a \code{clean()}
call. The extended \pkg{dataMaid} output document should have the
following modifications, relative to the standard \pkg{dataMaid} output:
\begin{itemize}
\item We want to add the new pre-check function, \code{isID}, to the already existing pre-checks.
\item We wish to change the plot type for all variables to the new mosaic plot.
\item We want the new summary function, \code{countZeros}, to be added to the summaries performed on all variable types but \code{Date} and \code{logical}.
\item We want the new check function, \code{identifyColon}, to be added to the checks performed on \code{character}, \code{factor} and \code{labelled} variables.
\end{itemize}
These options are specified as follows: %\hl{Jeg synes ikke testData egner sig her. Der er ingen koloner og CPRvar er et mærkeligt, internt navn, hvis ikke man kender danske CPR-numre. Langt de fleste variable har desuden ingen nuller. Måske skal der laves et nyt datasæt?}
\begin{Schunk}
\begin{Sinput}
> data("exampleData")
> clean(exampleData,
      # Add the new pre-check function
      preChecks = c("isKey", "isEmpty", "isID"),
      # Change the visual
      allVisuals = "mosaicVisual",
      # Add the new summary for specified data types
      characterSummaries = c(defaultCharacterSummaries(), "countZeros"),
      factorSummaries = c(defaultFactorSummaries(), "countZeros"),
      labelledSummaries = c(defaultLabelledSummaries(), "countZeros"),
      numericSummaries = c(defaultNumericSummaries(), "countZeros"),
      integerSummaries = c(defaultIntegerSummaries(), "countZeros"),
      # Add the new check for specific data types
      characterChecks = c(defaultCharacterChecks(), "identifyColons"),
      factorChecks = c(defaultFactorChecks(), "identifyColons"),
      labelledCheck = c(defaultLabelledChecks(), "identifyColons"))
\end{Sinput}
\end{Schunk}
The outputted document is found in Appendix \ref{sec:appendix2}.

@@

\subsubsection{Handling the figures}
Though figures give a nice overview of each variable, they are also
quite heavy objects in terms of memory allocation. Therefore, it might
be beneficial to not include figures in the \pkg{dataMaid} outputs for
very large datasets. This is controlled via the \code{mode} argument:
\begin{Schunk}
\begin{Sinput}
> clean(toyData, mode = c("summarize", "check"))
\end{Sinput}
\end{Schunk}
If figures are indeed needed, a different approach is to choose the
less memory heavy standard \proglang{R} figure style instead of the
\pkg{ggplot2} figures that are the default option in \code{clean()}. This
can be done by setting the \code{allVisuals = "basicVisual"} argument:
\begin{Schunk}
\begin{Sinput}
> clean(toyData, allVisuals = "basicVisual")
\end{Sinput}
\end{Schunk}
Of course, even less heavy plots might be achieved by writing new
\code{visualFunction}s, using the guidelines from section
\ref{sec:functionTemplates}. For instance, a future extension of
\pkg{dataMaid} might be the inclusion of ASCII plots, as
e.g., represented in the \proglang{R} package \pkg{txtplot}.

@@

\begin{Schunk}
\begin{Sinput}
> clean(toyData, render = FALSE, openResult = FALSE)
> render("dataMaid_toyData.Rmd", quiet = FALSE)
\end{Sinput}
\end{Schunk}
This also deals with the fact that \pkg{dataMaid} can produce
\pkg{rmarkdown} files that supersedes the upper size limit of code editors, e.g.,
RStudio, which currently has an editor file size limit of 5MB. %\hl{dette giver
%ingen mening uden en præcision? "upper size limit" af hvad? Se udkommenteret
%kommentar.}
% of RStudio,
%which is currently \hl{find number} GBs (using RStudio version
%1.0.44).
%\hl{Is this maybe too editor specific? On the other hand, a
%  lot of people do use RStudio...}.

\subsection[Using dataMaid for problem flagging]{Using \pkg{dataMaid} for problem flagging}
If the data is large, but memory issues and computation time are less
of an issue than the human time it takes to look through the data
cleaning document, a viable solution might be not to include all
information about all variables. Or even for more reasonably sized
datasets, sometimes a brief overview of the most pressing issues can
be useful. This can be achieved by using the \code{onlyProblematic}
argument in \code{clean()}. By specifying \code{onlyProblematic = TRUE},
only variables that raise a flag in the checking steps will be
summarized and visualized. But perhaps we are not even interested in
obtaining general information about these variables, but only in
getting a quick overview of the problems they might have. This can be
done by also controlling the \code{mode} argument:
\begin{Schunk}
\begin{Sinput}
> clean(toyData, onlyProblematic = TRUE, mode = c("check"))
\end{Sinput}
\end{Schunk}
Now only the checking results are printed, and only for variables where problems were identified. An even more minimal output can be generated by also leaving out the checking results --- then \code{clean()} essentially just produces a list of the variable names that should be investigated further:
\begin{Schunk}
\begin{Sinput}
> clean(toyData, onlyProblematic = TRUE, mode = NULL)
\end{Sinput}
\end{Schunk}
Of course, this can also be done without generating an overview
document, by direct, interactive use of the \code{check()} function. When
called on a \code{data.frame}, this function produces a list (of
variables) of lists (of checks) of lists (or rather,
\code{checkResult}s). Thus, the overall problem status of each variable
can easily be unravelled using the list manipulation function
\code{sapply()}:
\begin{Schunk}
\begin{Sinput}
> toyChecks <- check(toyData)
> foo <- function(x) {
>   any(sapply(x, function(y) y[["problem"]]))
> }
> sapply(toyChecks, foo)
\end{Sinput}
\begin{Soutput}
 var1  var2  var3  var4  var5  var6
 TRUE  TRUE  TRUE  TRUE  TRUE FALSE
\end{Soutput}
\end{Schunk}
and we find that only a single variable in \code{toyData}, \code{var6} (for which all
observations have the value \code{"Irrelevant"}), is problem-free.

@@

In the this example, the \code{dataMaid\_toyData.Rmd} file could have been created as follows:
\begin{Schunk}
\begin{Sinput}
> clean(toyData, standAlone = FALSE)
\end{Sinput}
\end{Schunk}
and the more minimal, html-style \pkg{rmarkdown} file described above can be produced using
\begin{Schunk}
\begin{Sinput}
> clean(toyData, standAlone = FALSE, output = "html", twoCols = FALSE)
\end{Sinput}
\end{Schunk}
Note that with the latter option, no special YAML preamble is needed in the \pkg{rmarkdown} document.
% \hl{don't end on code.}

@@

While the the current release is stable, the authors have an interest
in further developing the functionality of \pkg{dataMaid} by providing
more summary, visual, and check function as part of the default
package. We are also currently considering adding options to handle
repeated measurement, where the visualizations --- in particular ---
might be improved by visualizing measurements over time. In addition,
an online \pkg{shiny} application where users that are not
\proglang{R}-savvy can upload their data and get a data cleaning
document is currently planned.

\nocite{R}
\nocite{shiny}
\nocite{dataMaid}
\nocite{rmarkdown}
\nocite{ggplot2}
\nocite{plyr}
\nocite{data.table}
\nocite{validate}
\nocite{editrules}
\nocite{janitor}
\nocite{DataCombine}
\nocite{txtplot}

>>>>

<<echo=FALSE>>=
# options(prompt = "R> ", width=77, continue = "+  ")
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
@

@@

Online tools such as OpenRefine (\url{http://openrefine.org/}) and
\proglang{R}-packages such as \pkg{plyr} \citep{plyr}, and
\pkg{data.table} \citep{data.table} have
made data wrangling a lot easier, but only a handful of packages such
as \pkg{editrules} \citep{editrules}, \pkg{validate} \citep{validate},
\pkg{DataCombine} \citep{DataCombine}, and
\pkg{janitor} \citep{janitor} attempt to implement systematic, reproducible data
cleaning.  These packages use different approaches for data cleaning:
\pkg{editrules} and \pkg{validate} provide frameworks for setting up
and checking constraints on the variables, while \pkg{DataCombine} and
\pkg{janitor} both provide a few functions for identifying problems
(e.g, duplicates, dates coded as numbers, etc.) in data.

While these tools attempt to alleviate the ubiquitous ad hoc approach
to data cleaning, they are primarily intended for the data savvy users
and less so for the general researcher with a knowledge about a specific
field and the context of the available data. The \pkg{dataMaid}
package \citep{dataMaid}
tries to address this by providing a framework that both allows for
extendable, systematic, reproducible data cleaning, and summarizing
findings for researchers from other fields such that they can act as
human experts when tracking down potential errors.

@@

The \pkg{dataMaid} approach to data cleaning and -quality assessment is
governed by two fundamental paradigms. First of all, there is no need
for data cleaning to be an ad hoc procedure. Often, we have a very
clear idea of what flags are raisable in a given dataset before we
look at it, as we were the ones to produce it in the first place. This
means that data cleaning can easily be a well-documented,
well-specified procedure. In order to aid this paradigm, \pkg{dataMaid}
provides easy-to-use, automated tools for data quality assessment in
\proglang{R} \citep{R} on which data cleaning decisions can be made. This quality
assessment is presented in an auto-generated overview document,
readable by data analysts and field experts alike, thereby also
contributing to an inter-field dialogue about the data at
hand. Oftentimes, e.g., distinguishing between faulty codings of a
numeric value and unusual, but correct, values requires
problem-specific expertise that might not be held by the data
analyst. Hopefully, having easy access to data descriptions through
\pkg{dataMaid} will help this necessary knowledge sharing.

@@

The final rendering of the generated markdown file is controlled by
the \code{render} and \code{openResult} arguments, which both default to
\code{TRUE}. \code{render} determines if the \proglang{R} markdown file
produced should be rendered using the \pkg{rmarkdown} \citep{rmarkdown} package and
\code{openResult} decides whether the output html or pdf file should be
opened. The following command produces an \proglang{R} markdown file
containing the data cleaning summary information, but without
rendering nor opening the markdown file:

@@

Note that the data visualizations in the cleaning summary are also
controllable, though only a single function can be provided for all
variable types. If for instance, we wish to change the visualizations
from the default \pkg{ggplot2} \citep{ggplot2} style histograms and barplots to base
\proglang{R} histograms and barplots, we type

@@

\begin{table}[htbp]
\footnotesize
\bgroup
\def\arraystretch{1.8}%
\begin{tabular}{L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}}
& \code{summaryFunction} & \code{visualFunction} & \code{checkFunction} \\
\hline
\vspace{0pt} Input (required) & \code{v} - a variable vector \newline \code{...}  &  \code{v} - a variable vector \newline \code{vnam} - the variable name (as character string) \newline \code{doEval} - a logical (\code{TRUE}/\code{FALSE}) controlling the output type of the function & \code{v} - a variable vector \newline \code{nMax} - an integer (or \code{Inf}), controlling how many problematic values are printed, if relevant \newline \code{...}  \\
Input (optional) &  \code{maxDecimals} - number of decimals printed in outputted numerical values.  & - &  \code{maxDecimals}  - number of decimals printed in outputted numerical values.  \\
Purpose & Describe some aspect of the variable, e.g., a central value, its dispersion or level of missingness. & Produce a distribution plot. & Check a variable for a specific issue and, if relevant, identify the values in the variable that cause the issue. \\
Output (required) & A list with entries: \newline \code{\$feature} - a label for the summary value (as character string); \newline \code{\$result} - the result of the summary (as character string) & A character string with \proglang{R} code for producing a plot. This code should be standalone, i.e., should include the data if necessary. & A list with entries: \newline \code{\$problem} - a logical identifying whether an issue was found; \newline \code{\$message} - a character string (possibly empty) decribing the issue that was found, properly escaped and ready for use in \code{rmarkdown} \\
Output (recommended) & A \code{summaryResult} object, i.e., an attributed list with entries \code{\$feature}, \code{\$result} and \code{\$value}, the latter being the values from \code{\$result} in their original format). & \bgroup
\def\arraystretch{1.8}%
\hspace*{-.2cm}\begin{tabular}[t]{p{0.45\linewidth} |
                 p{0.45\linewidth}} \vspace*{-1cm}\raggedright If
                 \code{doEval} is \code{TRUE}: \newline A plot that is
                 opened by the graphic device in \proglang{R}. &
                                                          \vspace*{-1cm}\raggedright If \code{doEval} is \code{FALSE}: \newline   A text string with \proglang{R} code, as described above. \end{tabular} \egroup  & A \code{checkResult} object, i.e., an attributed list with entries \code{\$problem}, \code{\$message} and \code{\$problemValues}, the latter being either \code{NULL} or the problem causing values, as they were found in \code{v}, whichever is relevant. \\
Tools available for producing the function & \code{summaryResult()} & --- & \code{messageGenerator()} \newline \code{checkResult()}
\end{tabular}
\egroup
\caption{Reference information for creating new functions to be used
  as part of the \mbox{summarize-,} visualize-, and check steps. The three
  columns correspond to each of the three function types. %\hl{I think the formatting here is still sort of chaotic. Maybe include horizontal lines everywhere? Is it allowed in JSS? It's impossible to read what's in which row right now.}
}
\label{table.functionTypes}
\end{table}

@@

This function can now be called directly or used in \code{clean()}, as will
presented in an example below. Depending on the \code{doEval} argument, either a text string
with code or a plot is produced. The plot resulting from the following
call is found in Figure~\ref{fig:mosaicPlot}:

<<fig=FALSE>>=
mosaicVisual(toyData$var1, "variable 1", doEval = TRUE)
@

@@

As with \code{mosaicVisual()}, we change the class of this function in
order to make it appear in \code{allSummaryFunctions()} calls. But now we
wish to emphasize that the function is not intended to be called on
all variable types, as zeros have different roles in \code{Date}s and in
\code{logical} variables:

XXX indentation
<<>>=
countZeros <- summaryFunction(countZeros,
  description = "Count number of zeros",
  classes = c("character", "factor", "integer",
  "labelled", "numeric"))
@

\subsubsection{identifyColons --- a new checkFunction with problem values}
The last function mentioned above is \code{identifyColons()}. We define
it using the helper function \code{messageGenerator()} to obtain a properly
escaped message, and we use \code{checkResult} to make its output print
neatly

XXX comments

<<>>=
identifyColons <- function(v, nMax = Inf, ... ) {
  v <- unique(na.omit(v))
  problemMessage <- "Note: The following values include colons:"
  problem <- FALSE
  problemValues <- NULL

@@

identifyColons <- checkFunction(identifyColons,
    description = "Identify colons surrounded by alphanumeric characters",
    classes = c("character", "factor", "labelled"))
@

As with the previous two functions, we also change its class. Note,
however, that for \code{checkFunction}s, the function description will
appear in the document produced by \code{clean()} (in the \textit{Data
  cleaning summary} section of the output), so now this is not only
done for the sake of the \code{allCheckFunctions()} output.

\subsubsection{Calling the new summarize/visualize/check functions from clean()}
Now, we are ready to use these new functions in a \code{clean()}
call. The extended \pkg{dataMaid} output document should have the
following modifications, relative to the standard \pkg{dataMaid} output:
\begin{itemize}
\item We want to add the new pre-check function, \code{isID}, to the already existing pre-checks.
\item We wish to change the plot type for all variables to the new mosaic plot.
\item We want the new summary function, \code{countZeros}, to be added to the summaries performed on all variable types but \code{Date} and \code{logical}.
\item We want the new check function, \code{identifyColon}, to be added to the checks performed on \code{character}, \code{factor} and \code{labelled} variables.
\end{itemize}
These options are specified as follows: %\hl{Jeg synes ikke testData
                                %egner sig her. Der er ingen koloner
                                %og CPRvar er et mærkeligt, internt
                                %navn, hvis ikke man kender danske
                                %CPR-numre. Langt de fleste variable
                                %har desuden ingen nuller. Måske skal
                                %der laves et nyt datasæt?}
<<eval=FALSE>>=
data("exampleData")
clean(exampleData,
  preChecks = c("isKey", "isEmpty", "isID"),
  allVisuals = "mosaicVisual",
  characterSummaries = c(defaultCharacterSummaries(), "countZeros"),
  factorSummaries = c(defaultFactorSummaries(), "countZeros"),
  labelledSummaries = c(defaultLabelledSummaries(), "countZeros"),
  numericSummaries = c(defaultNumericSummaries(), "countZeros"),
  integerSummaries = c(defaultIntegerSummaries(), "countZeros"),
  characterChecks = c(defaultCharacterChecks(), "identifyColons"),
  factorChecks = c(defaultFactorChecks(), "identifyColons"),
  labelledCheck = c(defaultLabelledChecks(), "identifyColons"))
@

The outputted document is found in Appendix \ref{sec:appendix2}.

@@

\subsubsection{Handling the figures}
Though figures give a nice overview of each variable, they are also
quite heavy objects in terms of memory allocation. Therefore, it might
be beneficial to not include figures in the \pkg{dataMaid} outputs for
very large datasets. This is controlled via the \code{mode} argument:

<<eval=FALSE>>=
clean(toyData, mode = c("summarize", "check"))
@

If figures are indeed needed, a different approach is to choose the
less memory heavy standard \proglang{R} figure style instead of the
\pkg{ggplot2} figures that are the default option in \code{clean()}. This
can be done by setting the \code{allVisuals = "basicVisual"} argument:

Of course, even less heavy plots might be achieved by writing new
\code{visualFunction}s, using the guidelines from section
\ref{sec:functionTemplates}. For instance, a future extension of
\pkg{dataMaid} might be the inclusion of ASCII plots, as
e.g., represented in the \proglang{R} package \pkg{txtplot} \citep{txtplot}.

@@

<<eval=FALSE>>=
clean(toyData, render = FALSE, openResult = FALSE)
render("dataMaid_toyData.Rmd", quiet = FALSE)
@

This also deals with the fact that \pkg{dataMaid} can produce
\pkg{rmarkdown} files that supersedes the upper size limit of code editors, e.g.,
RStudio, which currently has an editor file size limit of 5MB. %\hl{dette giver
%ingen mening uden en præcision? "upper size limit" af hvad? Se udkommenteret
%kommentar.}
% of RStudio,
%which is currently \hl{find number} GBs (using RStudio version
%1.0.44).
%\hl{Is this maybe too editor specific? On the other hand, a
%  lot of people do use RStudio...}.

\subsection[Using dataMaid for problem flagging]{Using \pkg{dataMaid} for problem flagging}
If the data is large, but memory issues and computation time are less
of an issue than the human time it takes to look through the data
cleaning document, a viable solution might be not to include all
information about all variables. Or even for more reasonably sized
datasets, sometimes a brief overview of the most pressing issues can
be useful. This can be achieved by using the \code{onlyProblematic}
argument in \code{clean()}. By specifying \code{onlyProblematic = TRUE},
only variables that raise a flag in the checking steps will be
summarized and visualized. But perhaps we are not even interested in
obtaining general information about these variables, but only in
getting a quick overview of the problems they might have. This can be
done by also controlling the \code{mode} argument:

<<eval=FALSE>>=
clean(toyData, onlyProblematic = TRUE, mode = c("check"))
@

Now only the checking results are printed, and only for variables
where problems were identified. An even more minimal output can be
generated by also leaving out the checking results --- then
\code{clean()} essentially just produces a list of the variable names
that should be investigated further:

<<eval=FALSE>>=
clean(toyData, onlyProblematic = TRUE, mode = NULL)
@

Of course, this can also be done without generating an overview
document, by direct, interactive use of the \code{check()} function. When
called on a \code{data.frame}, this function produces a list (of
variables) of lists (of checks) of lists (or rather,
\code{checkResult}s). Thus, the overall problem status of each variable
can easily be unravelled using the list manipulation function
\code{sapply()}:

<<>>=
toyChecks <- check(toyData)
foo <- function(x) {
  any(sapply(x, function(y) y[["problem"]]))
}
sapply(toyChecks, foo)
@

and we find that only a single variable in \code{toyData}, \code{var6} (for which all
observations have the value \code{"Irrelevant"}), is problem-free.

@@

In the this example, the \code{dataMaid\_toyData.Rmd} file could have
been created as follows:

<<eval=FALSE>>=
clean(toyData, standAlone = FALSE)
@

and the more minimal, html-style \pkg{rmarkdown} file described above can be produced using

<<eval=FALSE>>=
clean(toyData, standAlone = FALSE, output = "html", twoCols = FALSE)
@

Note that with the latter option, no special YAML preamble is needed in the \pkg{rmarkdown} document.
% \hl{don't end on code.}

@@

While the the current release is stable, the authors have an interest
in further developing the functionality of \pkg{dataMaid} by providing
more summary, visual, and check function as part of the default
package. We are also currently considering adding options to handle
repeated measurement, where the visualizations --- in particular ---
might be improved by visualizing measurements over time. In addition,
an online \pkg{shiny} \citep{shiny} application where users that are not
\proglang{R}-savvy can upload their data and get a data cleaning
document is currently planned.

% \nocite{R}
%\nocite{shiny}
%\nocite{dataMaid}
%\nocite{rmarkdown}
%\nocite{ggplot2}
%\nocite{plyr}
%\nocite{data.table}
%\nocite{validate}
%\nocite{editrules}
%\nocite{janitor}
%\nocite{DataCombine}
%\nocite{txtplot}
